### 2023-08-31 sections/1-introduction.tex

\section{Introduction}
\label{sec:introduction}
Often dubbed "the Big Data explosion," rapidly advancing technologies have enabled researchers to collect complex data at an unprecedented scale. 
This is especially true in the field of astronomy. 
The Sloan Digital Sky Survey (SDSS), for example, in its APOGEE dataset, has collected near-infrared spectra of approximately 430,000 stars in approximately 9,000 dimensions~\cite{alam2015eleventh}. 
Their MaNGA dataset also contains spectral measurements across the face of each of about 10,000 galaxies across 2,700 deg$^2$. [add sentence about big ANN competition]

Biology has also seen an influx of large, high-dimensional datasets. Silva 18S contains ribosomal DNA sequences of approximately 2.25 
million genomes with aligned length of 50,000 letters.

% (TODO: Do we need to talk about NR99; it seems like more of the same but 
% with removed redundancy; not sure if that is helpful to us...) 
% No we don't... I'll explain why if you want.

In many fields, the rate of data collection outpaces the rate of computing performance improvements predicted by Moore's Law~\cite{moore1965cramming}, indicating 
that computer architecture improvements will not "catch up" to our computational needs in the near future. As a result, analysis of these large datasets 
requires better algorithms.

Many researchers are especially interested in performing similarity search on these datasets. 
Similarity search enables a variety of applications, including recommendation and classification systems. 
Additionally, with the rise of large language models such as GPT-3~\cite{brown2020language}, similarity search can become a useful tool for 
understanding and verifying the validity of model outputs. 
As the sizes and dimensionalities of datasets have grown, however, efficient and accurate similarity search has become extremely challenging; 
even state-of-the-art algorithms exhibit a steep recall versus speed tradeoff~\cite{ishaq2019clustered}.

@@

Recent approaches to tackling the exponential growth of data include locality-sensitive hashing~\cite{bingham2001random}, 
clever indexing techniques such as the FM Index~\cite{ferragina2005indexing}, and entropy-scaling search~\cite{yu2015entropy, ishaq2019clustered}. 
Entropy-scaling search is a paradigm for similarity search that exploits the geometric and topological structure inherent in big datasets.
Importantly, as suggested by their name, entropy-scaling search algorithms have asymptotic complexity that scales with geometric properties (such as the 
\emph{metric entropy} and \emph{local fractal dimension}, defined in Section 2) of a dataset,
rather than its cardinality. In 2019, we introduced CHESS (Clustered Hierarchical Entropy-Scaling Search), which extended entropy-scaling $\rho$-nearest 
neighbors search to a hierarchical clustering approach. In this paper, we introduce CAKES (CLAM-Accelerated $K$-nearest-neighbor 
Entropy-scaling Search). CLAM (Clustered Learning of Approximate Manifolds), developed to allow ``manifold mapping'' for anomaly detection~\cite{ishaq2021clustered}, is a refinement of the CHESS clustering algorithm. 
Using the cluster tree constructed by CLAM, CAKES builds upon the capabilities of CHESS to perform 
$k$-nearest neighbors search in addition to $\rho$-nearest neighbors search (also known as ranged nearest neighbors search).

>>>>

\section{Introduction}
\label{sec:introduction}
Often dubbed "the Big Data explosion," rapidly advancing technologies have enabled researchers to collect complex data at an unprecedented scale. 
This is especially true in the field of astronomy. 
The Sloan Digital Sky Survey (SDSS)~\cite{blanton2017sdss}, for example, in its APOGEE~\cite{alam2015eleventh} dataset, has collected near-infrared spectra of approximately 430,000 stars in approximately 9,000 dimensions. 
Their MaNGA dataset also contains spectral measurements across the face of each of about 10,000 galaxies across 2,700 deg$^2$. [add sentence about big ANN competition]

Biology has also seen an influx of large, high-dimensional datasets. Silva 18S~\cite{10.1093/nar/gks1219} contains ribosomal DNA sequences of approximately 2.25 
million genomes with aligned length of 50,000 letters.

More recently, Big-ANN launched NeurIPS'23: Billion Scale Approximate Nearest Neighbor Search Challenge, 
which aims to encourage the development of indexing data structures and search algorithms for practical variants of Approximate Nearest Neighbor (ANN) or Vector search.
This challenge tests submissions' performances with the YFCC 100M~\cite{johnson2019billion} dataset, the Yandex Text-to-Image 10M~\cite{diskann-github}, cross-modal dataset, 
the MSMARCO passage retrieval dataset~\cite{Bruch2023AnAA}, and a 10M slice of the MS Turing data set~\cite{Singh2021FreshDiskANNAF}. These datasets 
are 192, 200, 30,000, and 100 dimensional respectively.

In many fields, the rate of data collection outpaces the rate of computing performance improvements predicted by Moore's Law~\cite{brescia2012extracting}, indicating 
that computer architecture improvements will not "catch up" to our computational needs in the near future. As a result, analysis of these large datasets 
requires better algorithms.

Many researchers are especially interested in performing similarity search on these datasets. 
Similarity search enables a variety of applications, including recommendation and classification systems. 
Additionally, with the rise of large language models such as GPT-3~\cite{2020arXiv200514165B}, similarity search can become a useful tool for 
understanding and verifying the validity of model outputs. 
As the sizes and dimensionalities of datasets have grown, however, efficient and accurate similarity search has become extremely challenging; 
even state-of-the-art algorithms exhibit a steep recall versus speed tradeoff~\cite{ishaq2019clustered}.

@@

Recent approaches to tackling the exponential growth of data include locality-sensitive hashing~\cite{indyk1999sublinear}, 
clever indexing techniques such as the FM Index~\cite{simpson2010efficient}, and entropy-scaling search~\cite{yu2015entropy, ishaq2019clustered}. 
Entropy-scaling search is a paradigm for similarity search that exploits the geometric and topological structure inherent in big datasets.
Importantly, as suggested by their name, entropy-scaling search algorithms have asymptotic complexity that scales with geometric properties (such as the 
\emph{metric entropy} and \emph{local fractal dimension}, defined in Section 2) of a dataset,
rather than its cardinality. In 2019, we introduced CHESS (Clustered Hierarchical Entropy-Scaling Search), which extended entropy-scaling $\rho$-nearest 
neighbors search to a hierarchical clustering approach. In this paper, we introduce CAKES (CLAM-Accelerated $K$-nearest-neighbor 
Entropy-scaling Search). CLAM (Clustered Learning of Approximate Manifolds), developed to allow ``manifold mapping'' for anomaly detection~\cite{ishaq2021clustered}, is a refinement of the CHESS clustering algorithm. 
Using the cluster tree constructed by CLAM, CAKES builds upon the capabilities of CHESS to perform 
$k$-nearest neighbors search in addition to $\rho$-nearest neighbors search (also known as ranged nearest neighbors search).

### 2023-08-31 sections/5-discussion.tex

GPU-acceleration for the more computationally intensive distance functions like wasserstein.

Live-updates as new points are added to the dataset.
For example, build tree with a subsample of the full data and then add the rest of the instances to simulate live-updates.

compression

More distance functions. For example, Tanimoto Distance using maximal-common-subgraph for molecular structures.

>>>>

For future work, we plan to investigate compression by representing differences at each level of the
binary tree, particularly for string or genomic data, where all differences are discrete.
When we descend a level in the cluster tree, it takes fewer bits to encode each point in terms of the center. 
However, each descent of a level in the cluster tree increases the bit-cost of specifying a particular cluster. 
This presents the following tradeoff: \emph{How do we algorithmically determine the depth to which we should cluster a dataset to achieve optimal compression?} 
In our future work, we will explore which properties of the dataset and of clusters can inform this optimal depth.

We also plan to use CAKES with anomaly detection. Our anomaly detection tool, CHAODA~\cite{ishaq2021clustered}, uses 
a variety of methods in an ensemble to determine an anomaly score for a point. We would like to explore the effectiveness of 
distance to the $k$-th nearest neighbor as an additional method in our ensemble.

Additionally, we would like to implement additional distance functions to improve the genericity of CAKES. 
For example, we would like to implement Tanimoto Distance using maximal-common-subgraph for molecular structures. Ultimately, 
we would also like to implement Wasserstein distance for dimensions higher than 2. In order to make this possible, 
we plan to implement GPU-acceleration for the more computationally-intensive distance functions like Wasserstein.

Finally, we would like to add the ability to view live-updates as new points are added to a dataset. For example, 
we could build the cluster tree with a subsample of the full data set and then add the rest of the instances to 
simulate live updates.
